---
title: "Tree simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tree simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

## Simulate

Section 6.1.1: three equally likely motifs, with supports
$\{1,2,3\}$, $\{1,2,4\}$, and $\{1,5\}$. One seed; one $\tau_0$.

```{r simulate}
library(misoR)
seed <- 20260918L
tau_0 <- 0.7
set.seed(seed)
N <- M <- 1000L
K <- 5L
D <- 3L
supports <- list(c(1, 2, 3), c(1, 2, 4), c(1, 5))
F <- matrix(rgamma(K * M, shape = 0.1), K, M)
F <- F / rowSums(F)
z <- sample.int(3, N, replace = TRUE, prob = rep(1/3, 3))
L <- matrix(0, N, K)
for (s in 1:3) {
  i <- which(z == s)
  L[i, supports[[s]]] <- matrix(
    rgamma(length(i) * length(supports[[s]]), shape = 1, rate = 1/50),
    nrow = length(i))
}
Y <- Matrix::Matrix(matrix(rpois(N * M, as.vector(L %*% F)), N, M), sparse = TRUE)
```

## Fit three motifs

Estimate $F$ from the counts. Reuse this NMF start for every fit.

```{r initialize}
start <- miso_init(Y, K = K, D = D, S = 6, tau = tau_0,
                   seed = seed, nmf_max_iters = 1000)
initialize <- function(S) miso_init(
  Y, F = start$F, L = start$expected_loading, D = D, S = S,
  tau = tau_0, seed = seed)
```

Fit a descending $4\to3\to2$ path; inspect its $S=3$ fit first.

```{r fit-three}
path <- miso_fit_path(Y, S = 4:2, init = initialize(4), seed = seed,
                     warm_up_iters = 20, max_iters = 2000,
                     tol = 1e-9, min_iters = 50, patience = 10)
stopifnot(all(path$summary$converged))
fit3 <- path$fits[["S3"]]

# Match factor labels to truth for interpretation only.
cosine <- (fit3$F %*% t(F)) / outer(sqrt(rowSums(fit3$F^2)), sqrt(rowSums(F^2)))
matched <- max.col(cosine, ties.method = "first")
stopifnot(length(unique(matched)) == K)
rownames(fit3$F) <- paste0("F", matched)
plot(fit3, type = "loadings", factor_order = order(matched),
     sort_by = "dominant", main = "S = 3")

L_hat <- predict(fit3, "loadings")[, order(matched)]
profile <- rowsum(L_hat, fit3$z_hat)
profile <- profile / rowSums(profile)
knitr::kable(round(profile, 3), col.names = paste0("F", 1:K),
             caption = "Loading shares within each inferred motif.")
```

F1 is shared by all motifs; F2 joins F3 or F4; F5 forms the third branch.
The small remaining contributions are retained.

## Fit six motifs

```{r fit-six}
fit6 <- miso_fit(Y, S = 6, init = start, warm_up_iters = 20,
                 max_iters = 2000, tol = 1e-9, min_iters = 50, patience = 10)
stopifnot(fit6$converged)
mass <- sort(fit6$allocation_mass, decreasing = TRUE, index.return = TRUE)
barplot(mass$x, names.arg = paste0("M", mass$ix), ylim = c(0, 0.4),
        ylab = "Posterior membership mass", main = "S = 6", col = "steelblue")
knitr::kable(data.frame(Motif = mass$ix, Mass = round(mass$x, 4)))
```

The three smallest motifs together receive
`r sprintf("%.1f%%", 100 * sum(tail(mass$x, 3)))` of membership mass.
They are small, not empty. Mass is $N^{-1}\sum_i q(Z_i=s)$.

## Select the motif count

Compare the same path: $K,D$ and the prior are fixed; smaller models start from merges.

```{r select-S}
plot(path)
knitr::kable(path$summary[, c("S", "ELBO")], digits = 2)
best_S <- path$S[which.max(path$elbo)]
```

The largest attained ELBO selects **$S=`r best_S`$** on this single path.
